Moisture absorption and drying alter nonadditive litter flammability in a mixed conifer forest

### 2020-05-01 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup

\documentclass[letterpaper,12pt]{article}
\usepackage{caption}
\DeclareCaptionLabelFormat{adja-page}{\hrulefill\\#1 #2 \emph{(previous page)}}
\usepackage{jecol}
\usepackage{endfloat}

@@

      \item Litter mixtures exhibited nonadditivity in flame spread rate and in
    fuel consumption. This nonadditivity became more positive with fuel drying

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond readily
to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
with climate, with seasons, and with short term weather within a season.
Therefore, both the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat total transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with flame heights and area burned, and with total heat
transfer predicting soil impacts of fire and tree mortality. A shortcoming in
our current knowledge is our integration of specific species and plant traits
into fuel hazard assessment, fire modeling, and fire effects prediction.
Developing quantitative yet robust and general ways to integrate actual plant
communities into fuel assessment will require understanding the magnitude of
flammability variability across species in space and time and parameterizing
based on plant characteristics.

Species differ in their litter flammability \citep{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015} and fire regime can vary over fine spatial scales
in response to community composition and species-specific flammability
\citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter
packing, measured as litter-bed density or packing ratio, is an important
control on ignition and spread in models of surface fire \citep{Weber-1991,
  Balbi+Santoni+etal-1999, Morvan+Larini-2001}. Multiple studies have
demonstrated that litter bed density controls the flammability of dry litter
fuels in seasonally dry climates where higher litter density reduces flame
spread rate. Across these studies, there is increasing consensus that larger,
less densely packed litter beds increase oxygen availability and burn with
faster spread rates \citep{Scarff+Westoby-2006, Schwilk+Caprio-2011,
  Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015, Varner+Kane+etal-2015,
  Kreye+Varner+etal-2018}. The litter packing is driven by litter particle size
and shape, which is in turn influenced by leaf size, how leaves or branches are
abscised, and subsequent decomposition \citep{Fonda+Belanger+etal-1998,
  Stephens+Finney+etal-2004, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015, Zhao+Cornwell+etal-2016,
  Grootemaat+Wright+etal-2017, Prior+Murphy+etal-2018,
  Zhao+vanLogtestijn+etal-2019}.

@@

Differential drying is especially important when precipitation occurs
during the same season as ignitions. Although synoptic weather conditions that
create dry and wet lightning storms differ, the potential for precipitation and
lightning during the same overall period of thunderstorm activity often exists:
wet and dry convection days can occur under overlapping conditions,
\citep{Rorig+Ferguson-1999}. Moisture levels of fine fuels respond rapidly to
precipitation events during the fire season \citep{Estes+Knapp+etal-2012}.
Differential moisture content across time may contribute to the often high
spatial heterogeneity of surface fires \citep{Hille+Stephens-2005,
  Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to variation in fire regime
across landscapes \citep{Stephens-2001}.

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
can indicate different abilities to retain moisture. Leaf morphology,
especially leaf cuticle thickness, can influence water absorption
\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
moisture content because a larger total surface area will increase water
adsorption. Drying rate should also be influenced by leaf and litter bed traits
because surface-area-to-volume ratio and litter bulk density will influence air
flow through the litter bed. Therefore, we expect litter packing to effect fire
behavior in two ways: directly by its effect on oxygen availability for
combustion, and indirectly by its effect on drying rate. Both these pathways
are influenced by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds comprising larger leaves or leaves with
thin cuticles should favor moisture absorption. Conversely, large leaves lower
litter density and should favor drying, thus increasing the probability of
ignition and fire spread. The interaction of these processes, especially in
species mixtures, is poorly known.

@@

This study is the first to examine ecologically relevant drying rates and their
interaction with non additive moisture effects on flammability. We asked if
differences in litter moisture absorption and drying rates could change the
relative flammability of litter across species compared with flammability
measurements based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water absorption and rate of moisture loss and examined
how these traits vary in relation to specific leaf area and litter packing
density. Experimental burn trials of leaf litter at various moisture levels
determined how time since wetting influenced two axes of flammability
characterized by flame spread rate and total fuel consumed. To determine the
potential magnitude of mixture effects, we tested if the nonadditive effects on
reported in previous studies occur under higher moisture levels, and whether
fuel moisture effects strengthen or moderate any nonadditivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 1\,cm diameter that have fallen in the previous year (mostly
undecomposed or lightly decomposed litter representing fuels classified as
1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
mid-July) in 2012 from 21 separate sites across the study area. We collected
from a minimum of four populations per species with each population within a
species was 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2--4 individual trees at each site which were separated
from one another by at least 10\,m. The litter was collected 2\,m away from the
trunk to obtain a more uniform sample of leaf and fine branchlet litter,
because bark and heavier branches tend to fall closer to the tree trunk. The
top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
All litter was oven dried at $60^\circ$C to $< 5$\% fuel moisture by dry
weight, then weighed and divided into moisture trials. The collected litter was
stored in large paper bags and cardboard boxes in a lab storage space. We
minimized handling and moving the bagged litter to avoid mechanical damage and
changes to the particle size distributions. The moisture manipulations and
subsequent burning trials were conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water absorption and the rate
of drying for each species and for four distinct three-species mixtures. These
curves allowed us to determine the rate at which a species loses moisture,
approximated as exponential decay, and to determine maximum water content
following saturation and gravity draining. Samples from the eight species (six
replicates per species) and four mixtures (five replicates per mixture type) were
placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
sensor (Maxim Integrated) placed 1\,m above the samples. Litter baskets were
arranged in a completely randomized design. During the dry-down process three
subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to
assess moisture content. This was repeated until the final sample from a basket
was less than 10\% moisture. This experimental setup was designed to accelerate
drying compared with field conditions because the baskets were suspended in air that was
2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the
closest weather stations to the collection sites (Grant Grove and Lodegpole
stations, NOAA 1981-2010 normals,
https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

@@

Past work has demonstrated that green leaf length predicts litter packing and,
consequently, flame spread rate across these species. Litter particle size
distributions, however show that there is substantial variation in particle
sizes within a species (SI Fig.~1). There were differences in the maximum water
absorption (all $P < 0.0001$, Fig.~\ref{fig:drydown} and
Table~\ref{tab:drydown}). \emph{Q. kelloggii}, the only broadleaf, is the
species that absorbed the most moisture initially. \emph{Quercus}, like the
pines, had a rapid drying rate (Fig.~\ref{fig:drydown} and SI Fig.~2).
Calculating estimated marginal means following the linear mixed-effects model
discriminated the species into three different maximum water absorption groups:
1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and
3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P.
  jeffreyi} and \emph{P. ponderosa} (SI Fig.~2). There are also differences in
drying rates across the eight species and estimated mixed linear model slopes
discriminated species into three groups from lowest to most rapid: 1) \emph{C.
  decurrens}, 2) \emph{Abies spp.} and \emph{S. giganteum}, and 3) \emph{Q.
  kelloggii} and \emph{Pinus spp.} (SI Fig.~3).

@@

Across these eight species, specific leaf area had a positive association with
maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model
$P = 0.003$) and litter bulk density had a negative effect on drying rate
(Table~\ref{tab:mc_di_anova}, SI Fig~6, linear model $P = 0.029$).

\begin{table}
  \caption{Linear model results for maximum water absorption (\% moisture
    content on dry mass basis) and for estimated drying rate (hr$^{-1}$) as
    functions of specific leaf area and litter bulk density across eight
    species. ANOVA tables shown with degrees of freedom, sum of squares, mean
    squares, $F$, and $P$ for each effect. Model coefficients in SI Tables~2 and 3.}
  \label{tab:mc_di_anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.

\begin{tabular}{ccrrrrr}
  \toprule
Ind. variable & Effect & df & Sum of Squares & Mean squares & $F$ & $P$ \\ 
  \midrule
Moisture content & SLA & 1 & 250834.5 & 250834.51 & 130.13 & 0.000 \\ 
  & density & 1 & 538.5 & 538.54 & 0.28 & 0.625 \\ 
  & SLA:density & 1 & 5135.2 & 5135.24 & 2.66 & 0.178 \\ 
  & Residuals & 4 & 7710.4 & 1927.61 &  &  \\ 
  \midrule

Drying rate & density & 1 & $3.92 \times 10^{-4}$ & $3.92 \times 10^{-4}$ & 11.01 & 0.029 \\ 
 &  SLA & 1 & $1.88 \times 10^{-5}$ & $1.88 \times 10^{-5}$ & 0.53 & 0.507 \\ 
 &  density:SLA & 1 & $6.89 \times 10^{-6}$ & $6.89 \times 10^{-6}$ & 0.19 & 0.683 \\ 
 &  Residuals & 4 & $1.42 \times 10^{-4}$ & $3.56 \times 10^{-5}$ &  &  \\ 
   \bottomrule

\subsection*{Flammability across taxonomic groups}

Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}).
\emph{Quercus} had higher spread rates across all fuel moistures and had a
steeper decline in response to moisture than did the other groups (taxon by
time since wetting interaction $P = 0.022$, Table~\ref{tab:spreadrate_anova}).
Fuel moisture negatively influenced fuel consumption. Fuel consumption and the
rate of change with moisture content varied across the taxonomic groups
(Fig.~\ref{fig:flam_moist}, Table~\ref{tab:consume_anova}). Although the
overall ranking of groups was similar for fuel consumption and for flame spread
rate, there were different pattern across taxa in these two flammability
parameters. For spread rate, the decline in flammability with moisture was
shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:flam_moist}). For
fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
flammability with fuel moisture than did the other groups
(Fig.~\ref{fig:flam_moist}).

@@

Based on the estimated marginal
means for maximum water absorption and drying rate (SI Figs 3--4), we selected
four species for use in
the litter mixtures and examined all four three-species combinations of those
four. The four species were \emph{Abies concolor}, \emph{Calocedrus decurrens},
\emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.

@@

There was evidence for nonadditive effects in drying rate. Mixtures tended to
have lower water content than predicted from the mean of the constituent
species after wetting, but had higher water content than expected after long
drying (Wald test shows slope of mixed linear model
[Fig.~\ref{fig:mixture_obs_pred_mc}] less than 1, $P < 0.0001$). 
Mixture drying rates and maximum water absorption showed some relationship to
the dry down curves for individual constituent species (see SI Fig.~7). For
example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
Fig.~7) has the lowest maximum water absorption. Maximum water absorption and
drying rates differed across mixtures (linear model based on log of moisture
content, SI Tables~8 and 9).

@@

The taxa in this study displayed different patterns of both moisture absorption
and desorption, associated with different litter traits. In this sample of
species, those that have a greater absorptive capacity are also among the most
flammable in terms of spread rate when dry. Traits such as large leaves and
high SLA may affect both absorptive capacity and flammability with a
consequence that flammability rankings across species can change with time
since wetting. One shortcoming of past work in this system was that
flammability trials tested only dry fuels \citep{Magalhaes+Schwilk-2012}. We
therefore expected that rankings of species flammability might change under
higher moisture levels. For example, one presumption of ours prior to this
experiment was that the high flammability previously reported for \emph{Q.
  kelloggii} \citep{Magalhaes+Schwilk-2012} would disappear under higher fuel
moisture contents. This prediction was partly supported: \emph{Q. kelloggii}
fails to ignite until after 72 hours of drying in this study (Figs.
\ref{fig:drydown} and \ref{fig:drydown}). However, high moisture
content in this oak was unable to suppress ignition of mixtures to which it
contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}), indicating the importance of non-additive effects.

The dry down curves we measured demonstrate that after a rain event, the litter
of different species will exhibit different moisture contents over time as a
consequence of both differential initial absorption and differences in drying
rate. This leads to flammability varying both by taxon and time since wetting.
In these mixed conifer forests, fine fuels can vary in fuel moisture throughout
the fire season. For example, 10-hour fuels in northern California varied from
less than 10\% to over 80\% depending on precipitation events in a study
following fuel moisture throughout the fire season
\citep{Estes+Knapp+etal-2012}. Early summer prescribed burns at our study sites
in Sequoia National Park show evidence of moisture limiting fire behavior at
12-14\% fuel moisture (Knapp et al 2005).

>>>>

\documentclass[letterpaper,12pt]{article}
\usepackage{caption}
\DeclareCaptionLabelFormat{adja-page}{\hrulefill\\#1 #2 \emph{(previous page)}}
\usepackage{jecol}
\usepackage[notablist,nofiglist]{endfloat}
%\usepackage{color}
%\usepackage{soul}
%\renewcommand{\hl}[1]{#1}

@@

        \item  Litter mixtures exhibited nonadditivity in flame spread rate and in
    fuel consumption. This nonadditivity became more positive with fuel drying.

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. In particular, fine dead fuels respond readily to environmental
conditions such as temperature, relative humidity, and wind \citep{Nelson-2001,
  Kreye+Varner+etal-2018}. Moisture of litter fuels varies with climate, with
seasons, and with short term weather within a season. Therefore, both the rate
of drying and flammability as a function of moisture content are key parameters
that influence fuel flammability throughout the fire season. However, we lack
information on species-specific flammability responses to moisture dynamics. We
neither understand the relative importance of species differences in such
dynamics nor how these might interact to influence flammability in litter beds
comprising multiple species in forest communities.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In forests that experience surface fire,
litter properties influence fire behavior \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Varner+Kane+etal-2015}. In litter driven fire, the two
major axes of variation are described by measurements related to flame spread
rate and measurements capturing heat release or total heat transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with flame heights and area burned, and with heat transfer
predicting soil impacts of fire and tree mortality. A shortcoming in our
current knowledge is our integration of specific species and plant traits into
fuel hazard assessment, fire modeling, and fire effects prediction. Developing
quantitative yet robust and general ways to integrate actual plant communities
into fuel assessment will require understanding the magnitude of flammability
variability across species in space and time and parameterizing based on plant
characteristics.

Species differ in their litter flammability \citep{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015} and fire regime can vary over fine spatial scales
in response to such species-specific flammability \citep{Stephens-2001,
  Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter packing, measured as
litter-bed density or packing ratio, is an important control on ignition and
spread in models of surface fire \citep{Weber-1991, Balbi+Santoni+etal-1999,
  Morvan+Larini-2001}. Multiple studies have demonstrated that litter bed
density controls the flammability of dry litter fuels in seasonally dry
climates: higher litter density reduces flame spread rate. Across these
studies, there is increasing consensus that less densely packed litter beds
increase oxygen availability and burn with faster spread rates
\citep{Scarff+Westoby-2006, Schwilk+Caprio-2011, Magalhaes+Schwilk-2012,
  Cornwell+Elvira+etal-2015, Varner+Kane+etal-2015, Kreye+Varner+etal-2018}.
The litter packing is driven by litter particle size and shape, which is in
turn influenced by leaf size, how leaves or branches are abscised, and
subsequent decomposition \citep{Fonda+Belanger+etal-1998,
  Stephens+Finney+etal-2004, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015, Zhao+Cornwell+etal-2016,
  Grootemaat+Wright+etal-2017, Prior+Murphy+etal-2018,
  Zhao+vanLogtestijn+etal-2019}.

@@

Differential drying is especially important when precipitation occurs during
the same season as ignitions. Although synoptic weather conditions that create
dry and wet lightning storms differ, the potential for precipitation and
lightning during the same overall period of thunderstorm activity often exists:
wet and dry convection days can occur under overlapping conditions
\citep{Rorig+Ferguson-1999}. Moisture levels of fine fuels respond rapidly to
precipitation events during the fire season \citep{Estes+Knapp+etal-2012} and
this can lead to repeated wetting and drying periods during which lightning
ignition may occur. Such differential moisture content across time may
contribute to the often high spatial heterogeneity of surface fires
\citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to
variation in fire regime across landscapes \citep{Stephens-2001}.

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
can indicate different abilities to retain moisture. Leaf morphology,
especially leaf cuticle thickness, can influence water absorption
\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
moisture content because a larger total surface area will increase water
adsorption. Drying rate should also be influenced by leaf and litter bed traits
because surface-area-to-volume ratio and litter bulk density will influence air
flow through the litter bed. Therefore, we expect litter packing to effect fire
behavior in two ways: directly by its effect on oxygen availability for
combustion, and indirectly by its effect on drying rate. Both these pathways
are influenced by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds comprising larger leaves or leaves with
thin cuticles should favor moisture absorption. Conversely, large leaves lower
litter density and should favor drying, thus increasing the probability of
ignition and fire spread. The interactions among these processes, especially in
species mixtures, are poorly known.

@@

This study is the first to examine ecologically relevant drying rates and their
interaction with non additive moisture effects on flammability. We asked if
differences in litter moisture absorption and drying rates change the
relative flammability of litter across species compared with flammability
measurements based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water absorption and rate of moisture loss and examined
how these traits vary in relation to specific leaf area and litter packing
density. Experimental burn trials of leaf litter at various moisture levels
determined how time since wetting influenced two axes of flammability
characterized by flame spread rate and total fuel consumed. To determine the
potential magnitude of mixture effects, we tested if the nonadditive effects on
reported in previous studies occur under higher moisture levels, and whether
fuel moisture effects strengthen or moderate any nonadditivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider ``litter'' the top layer of leaves and small
twigs less than 1\,cm diameter that have fallen in the previous year (mostly
undecomposed or lightly decomposed litter representing fuels classified as
1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
mid-July) in 2012 from 21 separate sites across the study area. We collected
from a minimum of four populations per species with each population within a
species was 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2--4 individual trees at each site which were separated
from one another by at least 10\,m. The litter was collected 2\,m away from the
trunk to obtain a more uniform sample of leaf and fine branchlet litter,
because bark and heavier branches tend to fall closer to the tree trunk. The
top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
All litter was oven dried at $60^\circ$C to $< 5$\% fuel moisture by dry
weight, then weighed and divided into moisture trials. The collected litter was
stored in large paper bags and cardboard boxes in a lab storage space. We
minimized handling and moving the bagged litter to avoid mechanical damage and
changes to the particle size distributions. The moisture manipulations and
subsequent burning trials were conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water absorption and the rate
of drying for each species and for four distinct three-species mixtures. These
curves allowed us to determine the rate at which a species loses moisture,
approximated as exponential decay, and to determine maximum water content
following saturation and gravity draining. Samples from the eight species (six
replicates per species) and four mixtures (five replicates per mixture type)
were placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
sensor (Maxim Integrated) placed 1\,m above the samples. Litter baskets were
arranged in a completely randomized design. During the dry-down process three
subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to
assess moisture content. This was repeated until the final sample from a basket
was less than 10\% moisture. This experimental setup was designed to accelerate
drying compared with field conditions because the baskets were suspended in air
that was 2.8 -- 5.3$^\circ$C warmer than the historical average of the warmest
summer month at the closest weather stations to the collection sites (Grant
Grove and Lodegpole stations, NOAA 1981-2010 temperature means,
https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

@@

Across these species there were significant differences in the maximum water
absorption and drying rate (all $P < 0.0001$, Fig.~\ref{fig:drydown},
Table~\ref{tab:drydown} and SI Figs.~1 -- 3). Calculating estimated marginal
means following the linear mixed-effects model discriminated the species into
three different maximum water absorption groups: 1) \emph{Q. kelloggii}; 2)
\emph{P. lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica},
\emph{S. giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P.
  ponderosa} (SI Fig.~2). Estimated mixed linear model slopes discriminated
species into three groups from lowest to most rapid drying rate: 1) \emph{C.
  decurrens}, 2) \emph{Abies spp.} and \emph{S. giganteum}, and 3) \emph{Q.
  kelloggii} and \emph{Pinus spp.} (SI Fig.~3).

@@

Across these eight species, specific leaf area had a positive association with
maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~4, linear model
$P = 0.003$) and litter bulk density had a negative effect on drying rate
(Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model $P = 0.029$). Past work
has demonstrated that green leaf length predicts litter packing and,
consequently, flame spread rate across these species. Litter particle size
distributions, however show that there is substantial variation in particle
sizes within a species (SI Fig.~6).

\begin{table}
  \caption{Linear model results for maximum water absorption (\% moisture
    content on dry mass basis) and for estimated drying rate (hr$^{-1}$) as
    functions of specific leaf area and litter bulk density across eight
    species. ANOVA tables shown with degrees of freedom, sum of squares, mean
    squares, $F$, and $P$ for each effect. Model coefficients in SI Tables~2 and 3.}
  \label{tab:mc_di_anova}
\centering
\begin{tabular}{ccrrrrr}
  \toprule
Ind. variable & Effect & df & Sum of Squares & Mean squares & $F$ & $P$ \\ 
  \midrule
Moisture content & SLA & 1 & 250834.5 & 250834.51 & 130.13 & 0.000 \\ 
                 & density & 1 & 538.5 & 538.54 & 0.28 & 0.625 \\ 
                 & SLA:density & 1 & 5135.2 & 5135.24 & 2.66 & 0.178 \\ 
                 & Residuals & 4 & 7710.4 & 1927.61 &  &  \\ 
  \midrule
Drying rate & density & 1 & $3.92 \times 10^{-4}$ & $3.92 \times 10^{-4}$ & 11.01 & 0.029 \\ 
            &  SLA & 1 & $1.88 \times 10^{-5}$ & $1.88 \times 10^{-5}$ & 0.53 & 0.507 \\ 
            &  density:SLA & 1 & $6.89 \times 10^{-6}$ & $6.89 \times 10^{-6}$ & 0.19 & 0.683 \\ 
            &  Residuals & 4 & $1.42 \times 10^{-4}$ & $3.56 \times 10^{-5}$ &  &  \\ 
  \bottomrule
\end{tabular}
\end{table}

\subsection*{Flammability}

Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}) and
maximum rates and the steepness of decline differed by taxonomic group
(Table~\ref{tab:spreadrate_anova}). Similarly, fuel moisture negatively
influenced fuel consumption (Fig.~\ref{fig:flam_moist},
Table~\ref{tab:consume_anova}). Although the overall ranking of groups was
similar for fuel consumption and for flame spread rate, there were different
pattern across taxa in these two flammability parameters. For spread rate, the
decline in flammability with moisture was steepest for \emph{Quercus} and
shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:flam_moist}). For
fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
flammability with fuel moisture than did the other groups
(Fig.~\ref{fig:flam_moist}).

@@

\subsection*{Nonadditivity in moisture dynamics and flammability}

Based on the estimated marginal means for maximum water absorption and drying
rate (SI Figs 2--3), we selected four species for use in the litter mixtures
and examined all three-species combinations of those four. The four species
were \emph{Abies concolor}, \emph{Calocedrus decurrens}, \emph{Pinus jeffreyi}
and \emph{Quercus kelloggii}.

@@

Drying rate exhibited nonadditive effects. Mixtures tended to have lower water
content than predicted from the mean of the constituent species after wetting,
but had higher water content than expected after long drying (Wald test shows
slope of mixed linear model less than 1, $P < 0.0001$,
Fig.~\ref{fig:mixture_obs_pred_mc}). Mixture drying rates and maximum water
absorption showed some relationship to the dry down curves for individual
constituent species (see SI Fig.~7). For example, the mixture that does not
contain \emph{Q. kelloggii} (AbCaPi in SI Fig.~7) has the lowest maximum water
absorption. Maximum water absorption and drying rates differed across mixtures
(linear model based on log of moisture content, SI Tables~8 and 9).

@@

The taxa in this study displayed different patterns of both moisture absorption
and desorption, associated with different litter traits. Those species that had
a greater absorptive capacity were also among the most flammable in terms of
spread rate when dry. Traits such as large leaves and high SLA may affect both
absorptive capacity and flammability with a consequence that flammability
rankings across species can change with time since wetting. One shortcoming of
past work in this system was that flammability trials tested only dry fuels
\citep{Magalhaes+Schwilk-2012}. We therefore expected that rankings of species
flammability might change under higher moisture levels. For example, one
presumption of ours prior to this experiment was that the high flammability
previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012}
would lessen significantly under higher fuel moisture. This prediction
was partly supported: \emph{Q. kelloggii} fails to ignite until after 72 hours
of drying in this study (Figs. \ref{fig:drydown} and \ref{fig:drydown}).
However, high moisture content in this oak was unable to suppress ignition of
mixtures to which it contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}), indicating the importance of non-additive
effects.

The dry down curves we measured demonstrate that after a rain event, the litter
of different species will exhibit different moisture contents over time as a
consequence of both differential initial absorption and differences in drying
rate. This leads to local litter flammability varying both by taxon and time
since wetting. In these mixed conifer forests, fine fuels can vary in fuel
moisture throughout the fire season. For example, 10-hour fuels in northern
California varied from less than 10\% to over 80\% depending on precipitation
events in a study following fuel moisture throughout the fire season
\citep{Estes+Knapp+etal-2012}. Early summer prescribed burns at our study sites
in Sequoia National Park show evidence of moisture limiting fire behavior at
12-14\% fuel moisture \citep{Knapp+Keeley+etal-2005}.

Our flammability trials demonstrate that both flame spread rate and fuel
consumption increased with time since wetting but that the rate of increase
varied across these taxa. Only the pines are able to ignite and support flame
spread immediately after wetting (Fig. \ref{fig:drydown}). In contrast,
\emph{Quercus} exhibited a threshold response to drying with no ignition prior
to 72 hours of drying but once ignition is possible, it had faster spread rates
than the other taxa. Our two Cupressaceae, which have dense litter but have low
maximum water absorption, had low flammability across all moisture levels. This
result is consistent with the low flammability found in an Australian
Cupressaceae that have small litter particles that pack densely
\citep{Scarff+Westoby-2006, Bowman+Haverkamp+etal-2018}. However, densely
packed litter is not a consistent trait across the family because
\citet{Cornwell+Elvira+etal-2015} found relatively high litter flammability and
low litter density for some Asian Cupressaceae.
